---
title: "Mini-project 2"
author: "Ankit Mathur & Nitesh Jaswal"
date: "3/17/2019"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(broom)
library(arm)
library(GGally)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```


```{r}
# Load Data
dfp_WTHH = read.csv("DFP_WTHH_release.csv")
names(dfp_WTHH)[1] <- "rowid"
dfp = dplyr::select(dfp_WTHH, rowid, house3, presvote16post, weight_DFP, M4A, GREENJOB, WEALTH, MARLEG, ICE, GUNS, POP_1, POP_2, POP_3, gender)
# Removing NAs or 7 (Did not vote) for presvote16post
dfp =  dfp[!(is.na(dfp$presvote16post) | dfp$presvote16post == 7),]
# summary(dfp)
n_row = nrow(dfp)
```

## Variable definitions:
**Issue variables**:  
Respondents were asked to give their support for the following programs on a 1–5 scale, where 1 means *strongly support* and 5 means *strongly oppose*.  
• **M4A**: Medicare for All  
• **GREENJOB**: A Green Jobs program  
• **WEALTH**: A tax on wealth over $100 million  
• **MARLEG**: Legalizing marijuana  
• **ICE**: Defunding Immigration and Customs Enforcement  
• **GUNS**: Gun control

**Voter groups**:  
• **Loyal Democrats**: People who voted for Hillary Clinton in 2016 and a Democratic House candidate in 2018.  
• **Loyal Republicans**: People who voted for Donald Trump in 2016 and a Republican House candidate in 2018.  
• **Swing voters**: All other people who voted in 2018. In addition, define the following two subsets of swing voters:  
  • **Switch to D**: People who didn’t vote for Hillary Clinton in 2016 but voted for a Democratic House candidate in 2018.  
  • **Switch to R**: People who didn’t vote for Donald Trump in 2016 but voted for a Republican House candidate in 2018.  
\pagebreak

```{r}
is_swing_voter <- function() {
  vect = as.numeric(rep.int(1, times = n_row))
  for(i in 1:n_row) {
    if(dfp$presvote16post[i] == 1 && dfp$house3[i] == 1)
      vect[i] = 2
    else if(dfp$presvote16post[i] == 2 && dfp$house3[i] == 2)
      vect[i] = 3
    else if(dfp$presvote16post[i] != 2 && dfp$house3[i] == 2)
      vect[i] = 4
    else if(dfp$presvote16post[i] != 1 && dfp$house3[i] == 1)
      vect[i] = 5
  }
  return(vect)
}
# 1 = Swing, 2 = Loyal Demo, 3 = Loyal Rep, 4 = Switch to R, 5 = Switch to D
dfp$is_swing = is_swing_voter()
dfp$is_swing = factor(dfp$is_swing, labels = c("Swing", "Loyal Democrat", "Loyal Republican", "Switch to R", "Switch to D"))

```

```{r message = FALSE, warning = FALSE}
dfp.q1 = dfp %>% subset(., subset = (.$is_swing == "Switch to R" | (.$is_swing == "Switch to D")), select = c("is_swing", "M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS", "weight_DFP"))

cb_palette2 = c("#E69F00", "#56B4E9")

dfp.q1 = gather(dfp.q1, key = "ISSUE", value = "RESPONSE", c("M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS"))
dfp.q1$RESPONSE = factor(dfp.q1$RESPONSE)
dfp.q1 = na.omit(dfp.q1)
```

## **Q1: How do Switch to D and Switch to R voters differ on the issue variables?**

```{r}
# ggplot(dfp.q1[dfp.q1$RESPONSE != 6,], aes(x = RESPONSE, color = is_swing, fill = is_swing)) + stat_count((aes(weight = weight_DFP))) + facet_wrap(~ISSUE) + theme_bw() + guides(fill = guide_legend(title="Swing Voters"), color = FALSE) + scale_fill_manual(values = cb_palette2) + labs(title="Distribution of response on Issue Variables for Switch to D/R voters", caption = "Response Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") + ylab("Count") + xlab("Response")

ggplot(dfp.q1[dfp.q1$RESPONSE != 6,], aes(x = RESPONSE, color = is_swing, fill = is_swing)) +
  geom_bar(mapping = aes(x = RESPONSE, y = ..prop.., group = 1, weight=weight_DFP), stat = "count") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ISSUE + is_swing) + theme_bw() + 
  guides(fill = guide_legend(title="Swing Voters"), color = FALSE) +
  scale_fill_manual(values = cb_palette2) +
  labs(title="Distribution of response on Issue Variables for Switch to D/R voters", caption = "Response Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") +
  ylab("Proportion of voters (%)") + xlab("Response")
```

* From the plot above, it can be observed that the **Switch to D** voters have a strong opinion on all issues except the immigration issue (ICE). On the other hand, the **Switch to R** voters have a strong opinion on issues such as ICE, MARLEG and M4A.
* The **Switch to R** and **Switch to D** voters differ strongly on the Medicare for all (M4A) issue. While a majority of **Switch to R** voters seem to strongly oppose this issue, the **Switch to D** voters tend to support it.
* The **Switch to R** and **Switch to D** voters behave rather similarly on the issue of marijuana legalization (MARLEG) with both the cohorts showing support towards the legalization.
* It is interesting to note that while the **Switch to R** voter group does not seem to have a strong opinion on multiple issues (GREENJOB, GUNS, and WEALTH), the **Switch to D** cohort does demonstrate a strong opinion on all issues except immigration (ICE).

```{r}
WTHH_full = read_csv("DFP_WTHH_release.csv")
WTHH = WTHH_full[c("presvote16post", "house3", "weight_DFP", "M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS", "POP_1", "POP_2", "POP_3")]
```

```{r}
# WTHH$grp <- ifelse(WTHH$presvote16post==1 & WTHH$house3==1, "Loyal Democrats", ifelse(WTHH$presvote16post==2 & WTHH$house3==2, "Loyal Republicans", ifelse(!(WTHH$presvote16post %in% c(7,NA)), "Swing voters", NA)))
WTHH$grp <- ifelse(WTHH$presvote16post==1 & WTHH$house3==1, "Loyal Democrats", ifelse(WTHH$presvote16post==2 & WTHH$house3==2, "Loyal Republicans", ifelse(!is.na(WTHH$presvote16post), "Swing voters", NA)))

WTHH$swing_bhv <- ifelse(WTHH$grp=="Swing voters", ifelse(WTHH$house3==1, "Switch to D", ifelse(WTHH$house3==2, "Switch to R", "Others")), NA)

# cols <- c("presvote16post", "house3", "M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS", "POP_1", "POP_2", "POP_3", "grp", "swing_bhv")
# WTHH[cols] <- lapply(WTHH[cols], factor)
```
\pagebreak

```{r}
WTHH.long = gather(WTHH, key="issue", value="response",
                   c("M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS"))
cols <- c("issue", "response")
WTHH.long[cols] <- lapply(WTHH.long[cols], factor)
WTHH.long <- WTHH.long[c("issue", "response", "grp", "weight_DFP")]
WTHH.long <- WTHH.long[!is.na(WTHH.long$response) & !is.na(WTHH.long$grp),]
```

## **Q2: How do swing voters differ from loyal Democrats and loyal Republicans on the issue variables?**

```{r}
ggplot(WTHH.long[WTHH.long$response != 6 & (WTHH.long$grp=="Loyal Democrats" | WTHH.long$grp=="Swing voters"),], aes(x=response, color = grp, fill = grp)) +
  geom_bar(mapping = aes(x = response, y = ..prop.., group = 1, weight=weight_DFP), stat = "count") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~issue + grp) + scale_fill_manual(values=cb_palette2) +
  guides(fill = guide_legend(title="Swing Voters"), color = FALSE) +
  labs(fill="Voter group", caption="Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") +
  ggtitle("Distribution of Voter groups by Issue & Loyalty") +
  xlab("Response") + ylab("Number of voters") +
  theme_bw()

ggplot(WTHH.long[WTHH.long$response != 6 & (WTHH.long$grp=="Loyal Republicans" | WTHH.long$grp=="Swing voters"),], aes(x=response, color = grp, fill = grp)) +
  geom_bar(mapping = aes(x = response, y = ..prop.., group = 1, weight=weight_DFP), stat = "count") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~issue + grp) + scale_fill_manual(values=cb_palette2) +
  guides(fill = guide_legend(title="Swing Voters"), color = FALSE) +
  labs(fill="Voter group", caption="Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") +
  ggtitle("Distribution of Voter groups by Issue & Loyalty") +
  xlab("Response") + ylab("Number of voters") +
  theme_bw()
```
With the exception of immigration issue (ICE), **Swing voters** tend to share *similar* opinions with **Loyal Democrats** across all of the remaining five issues. Interestingly, both these cohorts voted in favor of all five issues with a majority of voters choosing the *Strongly support* option.  
Having said this, an equivalent number of **Swing voters** chose *Strongly support* and *Strongly oppose* on the Medicare for all (M4A) issue, thereby showcasing a slightly different opinion than the **Loyal Democrats**.  
As for the immigration issue, while **Loyal Democrats** do not seem to have any coherent opinion, both **Loyal Republicans** and **Swing voters** cohorts voted against the defunding of ICE with a majority of them choosing the *Strongly oppose* option.


# **Q3: What predicts being a swing voter?**
```{r}
WTHH$swing_voter = ifelse(WTHH$grp=="Swing voters", 1, ifelse(is.na(WTHH$grp), NA, 0))
WTHH.issue = WTHH[c("GREENJOB", "GUNS", "ICE", "M4A", "MARLEG", "WEALTH", "weight_DFP", "swing_voter")]
```

```{r}
WTHH.issue <- WTHH.issue[rowSums(WTHH.issue[,c(1,2,3,4,5,6)] == 6) == 0,]
WTHH.issue <- WTHH.issue[complete.cases(WTHH.issue),]
```

```{r}
WTHH.issue.long = gather(WTHH.issue, key="issue", value="response",
                   c("M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS"))
ggplot(WTHH.issue.long, aes(x=response, y=swing_voter)) + geom_point(alpha=0.4) +
  geom_jitter(height=0.1, width=0.25) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~issue) +
  labs(caption="Response Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") +
  ggtitle("Logistic regression fits for odds of Swing Voter by Issue") +
  xlab("Response") + ylab("Probability of Swing voter") +
  theme_bw()
```
The plots above showcase how the predicted probability of a registered voter being a swing voter varies with responses to each of the six issues. With the exception of GREENJOB and MARLEG, all the other fitted lines are fairly flat suggesting that we include just these two issues in our final model.  
However, just to be sure, we did explicitly fit a logistic regression model including all six issue variables without interaction and looked at the coefficients for each of these issues (see appendix). As expected, the coefficients for all issues, except GREENJOB and MARLEG, turned out to be close to zero.  
Hence, in our final model, we will focus only on these two issues. Furthermore, for the sake of simplicity more than anything else, we will avoid including any interactions between our predictors.  
Finally, let us fit a logistic regression model to predict swing voters using GREENJOB and MARLEG issues:
```{r}
WTHH.issue.logit <- glm(swing_voter ~ GREENJOB + MARLEG, weights=weight_DFP, family="quasibinomial", data=WTHH.issue)
display(WTHH.issue.logit)
```
Interpreting the model above, we get:
$$
\textrm{logit[P(Swing voter)]} =  - 1.75 + 0.25 \times \textrm{GREENJOB} - 0.19 \times \textrm{MARLEG}
$$
Using the "divide by 4" rule, we can say that:  
• As per our model, keeping the opinion on MARLEG constant and moving one level closer to opposing the GREENJOB issue *increases* the probability of swing voter by 6.25%  
• Similarly, keeping the opinion on GREENJOB constant and moving one level closer to opposing the MARLEG issue *decreases* the model probabilty by 4.75%  

Now, let's visualize this model:  

```{r}
issue.predict.grid = expand.grid(MARLEG = seq(1, 5, 1), GREENJOB = seq(1, 5, 1))
issue.pred = predict(WTHH.issue.logit, type = "response", newdata = issue.predict.grid)
issue.pred.df = data.frame(issue.predict.grid, swing.prob = as.vector(issue.pred))

ggplot(issue.pred.df, aes(x = MARLEG, y = swing.prob, group = factor(GREENJOB), color = factor(GREENJOB))) +
  geom_line(size=1) + xlab("MARLEG Response") + ylab("Probability of Swing voter") +
  labs(color = "GREENJOB \n Response", caption="Response Values: 1. Strongly support  2. Somewhat support  3. Neither support nor oppose \n  4. Somewhat oppose  5. Strongly oppose") +
  theme_bw() + scale_color_manual(values = cb_palette)
```


```{r}
# Raster 
ggplot(issue.pred.df, aes(x = MARLEG, y = GREENJOB, z = swing.prob)) + geom_raster(aes(fill = swing.prob))  + coord_fixed() + scale_fill_distiller(palette = "RdYlBu") + theme_bw() + labs(title="Predicted probability plot for a given response \non issue variables", fill = "Predicted \nProbability", caption="Response Values: 1. Strongly support  2. Somewhat support  \n3. Neither support nor oppose  4. Somewhat oppose  5. Strongly oppose" )
```

In the above plots, it can be clearly visualized that keeping the opinion on GREENJOB constant and moving towards opposing the MARLEG issue *decreases* the model probabilty of being a swing voter. Furthermore, for a given opinion on MARLEG, the more a voter opposes GREENJOB, higher are the chances of him/her being a swing voter.  

Moving on to *populism* variables, let's do a faceted plot (as before) to get a feel of how the odds of swing voter vary with each populism variable.

```{r}
WTHH.pop = WTHH[c("POP_1", "POP_2", "POP_3", "weight_DFP", "swing_voter")]
```

```{r}
WTHH.pop <- WTHH.pop[rowSums(WTHH.pop[,c(1,2,3)] == 6) == 0,]
WTHH.pop <- WTHH.pop[complete.cases(WTHH.pop),]
```

```{r}
WTHH.pop.long = gather(WTHH.pop, key="populism", value="response",
                   c("POP_1", "POP_2", "POP_3"))
ggplot(WTHH.pop.long, aes(x=response, y=swing_voter)) + geom_point(alpha=0.4) +
  geom_jitter(height=0.1, width=0.25) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~populism) +
  labs(caption="Response Values: 1. Strongly agree  2. Somewhat agree  3. Neither agree nor disagree \n  4. Somewhat disagree  5. Strongly disagree") +
  ggtitle("Logistic regression fits for odds of Swing Voter by Populism") +
  xlab("Response") + ylab("Probability of Swing voter") +
  theme_bw()
```

The slope of the fitted lines decrease as we move from *POP_1* to *POP_2* ultimately flattening out as we reach *POP_3*. As before, just for our satisfaction, we looked at the slopes for each of these populism variables by explicitly fitting a logistic regression model for predicting the swing voters and observed that the coefficients for *POP_2* and *POP_3* variables turned out to be close to zero (see appendix).  
Hence, let us fit our second model based on just the *POP_1* populism variable:
```{r}
WTHH.pop.logit <- glm(swing_voter ~ POP_1, weights=weight_DFP, family="quasibinomial", data=WTHH.pop)
display(WTHH.pop.logit)
```

We interpret the coefficients above as representing the odds of a registered voter being a swing voter as below:
$$
\textrm{logit[P(Swing voter)]} = - 0.79 - 0.28 \times{POP_1}
$$

Again, as per the "divide by 4" rule, we can say that an increment of one degree of disagreement with the sentiment shared by *POP_1* variable decreases the probability of swing voter by 7%.  

Visualizing this model we get:  

```{r}
pop.predict.grid = expand.grid(POP_1 = seq(1, 5, 1))
pop.pred = predict(WTHH.pop.logit, type = "response", newdata = pop.predict.grid)
pop.pred.df = data.frame(pop.predict.grid, swing.prob = as.vector(pop.pred))
ggplot(pop.pred.df, aes(x = POP_1, y = swing.prob)) +
  geom_line(size=1) + xlab("POP_1 Response") + ylab("Probability of Swing voter") +
  labs(caption="Response Values: 1. Strongly agree  2. Somewhat agree  3. Neither agree nor disagree \n  4. Somewhat disagree  5. Strongly disagree", title="Logistic regression fit for odds of swing voter") +
  theme_bw()
```

In order to better understand how these models perform, we have included the residual plots in the appendix.

\pagebreak

# Appendix

```{r}
WTHH.issue.logit1 <- glm(swing_voter ~ GREENJOB + GUNS + ICE + M4A + MARLEG + WEALTH, weights=weight_DFP, family="quasibinomial", data=WTHH.issue)
display(WTHH.issue.logit1)
```

```{r}
WTHH.issue.logit3 <- glm(swing_voter ~ GREENJOB * GUNS * ICE * M4A * MARLEG * WEALTH, weights=weight_DFP, family="quasibinomial", data=WTHH.issue)
display(WTHH.issue.logit3)
```

```{r}
WTHH.issue.logit.df <- WTHH.issue
WTHH.issue.logit.df$.fitted <- fitted.values(WTHH.issue.logit)
WTHH.issue.logit.df$.resid <- residuals(WTHH.issue.logit, type="response")
WTHH.issue.logit.df$.pred <- 1 * (WTHH.issue.logit.df$.fitted > max(WTHH.issue.logit.df$.fitted)/1)
Correct <- (WTHH.issue.logit.df$swing_voter == WTHH.issue.logit.df$.pred)
1 - (sum(Correct) / nrow(WTHH.issue.logit.df))
summary(factor(WTHH.issue.logit.df$swing_voter))
```

```{r residuals for issue}
ggplot(WTHH.issue.logit.df, aes(x=GREENJOB, y=.resid)) + geom_point(alpha=0.4) +
  geom_jitter(height=0.1, width=0.25) + geom_smooth(method = "loess", method.args = list(degree = 1)) +
  xlab("Response") + ylab("Residuals") + ggtitle("Residual plot for GREENJOB")
ggplot(WTHH.issue.logit.df, aes(x=MARLEG, y=.resid)) + geom_point(alpha=0.4) +
  geom_jitter(height=0.1, width=0.25) + geom_smooth(method = "loess", method.args = list(degree = 1)) +
  xlab("Response") + ylab("Residuals") + ggtitle("Residual plot for MARLEG")
```

```{r}
WTHH.pop.logit1 <- glm(swing_voter ~ POP_1 * POP_2 * POP_3, weights=weight_DFP, family="quasibinomial", data=WTHH.pop)
display(WTHH.pop.logit1)
```

```{r}
WTHH.pop.logit.df <- WTHH.pop
WTHH.pop.logit.df$.fitted <- fitted.values(WTHH.pop.logit)
WTHH.pop.logit.df$.resid <- residuals(WTHH.pop.logit, type="response")
WTHH.pop.logit.df$.pred <- 1 * (WTHH.pop.logit.df$.fitted > max(WTHH.pop.logit.df$.fitted)/1)
Correct <- (WTHH.pop.logit.df$swing_voter == WTHH.pop.logit.df$.pred)
1 - (sum(Correct) / nrow(WTHH.pop.logit.df))
summary(factor(WTHH.pop.logit.df$swing_voter))
```

```{r residuals for populism}
ggplot(WTHH.pop.logit.df, aes(x=POP_1, y=.resid)) + geom_point(alpha=0.4) +
  geom_jitter(height=0.1, width=0.25) + geom_smooth(method = "loess", method.args = list(degree = 1)) +
  xlab("Response") + ylab("Residuals") + ggtitle("Residual plot for POP_1")
```
